---
title: "Methods for comparing MAP among vegetation types and Biomes"
author: "Marcio"
date: "7/15/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
source("./script_rainfall_clean.R")
```


#### Here, we investigate differences in rainfall between the Amazon biome, the Cerrado biome, and three vegetation types within the Cerrado (i.e. Savanna, Gallery Forest, and Dry Forest).

### **Methods and Results**

The visual inspection of the residuals distribution show that it is normaly distributed, with the mean near zero (Figure 1). 

```{r, fig.cap="Figure 1: Residuals inspection.", warning=FALSE, message=FALSE}
lm(Rainfall ~ as.factor(Biome),
   data = data)$resid %>% hist
```

Firstly, we used the _gls_ function from the R package _nlme_ (Pinheiro et al., 2019) to fit a linear mixed-effects model. This model considered rainfall as the response variable, with the factors of interest being Biome (with two categories: Amazon and Cerrado) and Vegetation Type (with three categories: Savanna, Gallery Forest, and Dry Forest) (Table 1).

```{r, fig.cap="Table 1: GLS model summary output.", warning=FALSE, message=FALSE}
m1 <- nlme::gls(Rainfall ~ as.factor(Biome),
    data = data,
    weights = nlme::varIdent(form = ~ 1 | as.factor(Biome)),
    method = "ML", na.action = na.omit) 

summary(m1)$tTable
```

Next, we performed an analysis of variance (ANOVA) using the _anova_ function in R to assess the statistical significance of the main effects and the interaction between the factors.


```{r, fig.cap="Table 2: Anova output.", warning=FALSE, message=FALSE}
an_m <- anova(m1)
an_m %>% data.frame
```

ANOVA revealed statistically significant results (Table 2) and we proceeded with post hoc tests to conduct multiple comparisons between groups. For this purpose, we utilized the _emmeans_ function from the package _emmeans_ (Lenth, 2022) to compute adjusted means and confidence intervals for each combination of factors (Table 3).

```{r, fig.cap="Table 3: Statistics of MAP for all Biomes and vegetation types.", warning=FALSE, message=FALSE}
means <- emmeans::emmeans(m1, specs = list("Biome"))
means %>% data.frame
```

Subsequently, we conducted multiple comparison tests using the _contrast_ function, also from _emmeans_ package, with the argument _method = "pairwise"_. This test allowed us to identify significant differences between groups (or not) by comparing all factor levels, such as biomes and vegetation types. Adjusted p-values were considered to control the family-wise error rate.

```{r, fig.cap="Table 4: Pairwise comparisons of MAP among biomes and sampled vegetation types.", warning=FALSE, message=FALSE}
contrasts <- emmeans::contrast(means, method = "pairwise")

contrasts %>%
  data.frame %>%
  rename(Pair = "emmeans.of.Biome.contrast") %>% 
  rename(estimate = "emmeans.of.Biome.estimate") %>% 
  rename(se = "emmeans.of.Biome.SE") %>% 
  rename(df = "emmeans.of.Biome.df") %>% 
  rename(`t ratio` = "emmeans.of.Biome.t.ratio") %>% 
  rename(`p-vaue` = "emmeans.of.Biome.p.value") 
  
```


# Conclusion

We found no differences in MAP among our sampled vegetation types (Table 4). Furthermore, the MAP of our samples are not different from the MAP of both Biomes (Table 4).

Most of our samples are inside the bistability zone (1300-2100mm; Ciemer et al., 2019) (Figure 2). However they are near the lower limit with some dry forests overcoming the thresshold (Figure 2).

```{r, fig.cap="Figure 2: MAP distribution for Amazon and Cerrado Biomes, and for our sampled vegetation types.", warning=FALSE, message=FALSE}
data %>% ggplot(aes(x = factor(Biome), y = Rainfall))+
  geom_boxplot(data = data, outlier.shape = NA, show.legend = F, aes(fill = factor(Biome)))+
    scale_fill_manual(values = c("grey", "pink", "purple", "blue", "orange"))+
  ylim(c(0, 4000))+
  geom_vline(xintercept = 2.5, linetype = "dashed", size = 0.5, color = "grey")+
  geom_hline(yintercept = c(1300,2100), linetype = "dashed", color = "red")+
  labs(x = NULL, y = "Mean Annual Rainfall (mm)", title = NULL)+
  geom_text(aes(x = 3.5, y = 2000, label = "Bistability zone", color = "red", size = 11, ... = "sans"), show.legend = F)+
  geom_text(aes(x = 1.5, y = 3500, label = "Biomes", size = 11, ... = "sans"), show.legend = F)+
  geom_text(aes(x = 4, y = 3500, label = "Our samples", size = 11, ... = "sans"), show.legend = F)+
  geom_text(aes(x = 1, y = 3475, label = "a"))+
  geom_text(aes(x = 2, y = 2130, label = "b"))+
  geom_text(aes(x = 3, y = 1700, label = "b"))+
  geom_text(aes(x = 4, y = 1700, label = "b"))+
  geom_text(aes(x = 5, y = 1700, label = "b"))+
  theme(axis.text = element_text(size = 11),
        axis.title = element_text(size = 13))

```

This statistical approach facilitated a comprehensive and precise analysis of mean annual rainfall differences among biomes and vegetation types between and within Biomes and our sample sites.


### REFERENCES


Russell V. Lenth (2022). emmeans: Estimated Marginal Means, aka Least-Squares
Means. R package version 1.7.4-1. https://CRAN.R-project.org/package=emmeans

Pinheiro J, Bates D, DebRoy S, Sarkar D, R Core Team (2019). _nlme: Linear and
Nonlinear Mixed Effects Models_. R package version 3.1-140, <URL:
https://CRAN.R-project.org/package=nlme>.

